---
title: "Understanding World Population Dynamics"
subtitle: "Assignment 1 - PSYC593"
format:
  pdf: default
  html:
    self-contained: true
project:
  output-dir: ./doc/
editor: source
date: today
author: "Jordan M Navarro"
---

<!-- These problems come from Imai, K. & Webb Williams, N. (2018). Quantitative Social Science: An introduction -->
<!-- in tidyverse. Princeton University Press. -->

<!-- The code giving partial solutions by Víctor H Cervantes -->

Understanding population dynamics is important for many areas of
social science. We will calculate some basic demographic quantities
of births and deaths for the world's population from two time periods:
1950 to 1955 and 2005 to 2010. We will analyze the following CSV data
files - `Kenya.csv`, `Sweden.csv`, and `World.csv`. Each file
contains population data for Kenya, Sweden, and the world,
respectively. The table below presents the names and descriptions of
the variables in each data set.

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            Abbreviated country name
 
 `period`             Period during which data are collected
 
 `age`                Age group 
 
 `births`             Number of births in thousands (i.e., number of children 
                      born to women of the age group) 
 
 `deaths`             Number of deaths in thousands 
 
 `py.men`             Person-years for men in thousands 
 
 `py.women`           Person-years for women in thousands
--------------------------------------------------------------------------------

Source: United Nations, Department of Economic and Social Affairs, 
Population Division (2013). *World Population Prospects: The 2012 Revision, DVD Edition.*

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Load packages
library(stats)
library(rprojroot)
library(tidyverse)
```

```{r}
# Global variables

# Path variables

root_path <- rprojroot::find_root(has_dir("src")) # This is the root/working directory

# These are the subdirectories
code_path     <- file.path(root_path, "src") # Using the base R function file.path
docs_path     <- file.path(root_path, "doc")
raw_data_path <- file.path(root_path, "data", "raw_data")
figs_path     <- file.path(root_path, "results", "figures")
tables_path   <- file.path(root_path, "results", "tables")
```

```{r}
# Read data
world_data  <- read.csv(file = file.path(raw_data_path, "World.csv"))
kenya_data  <- read.csv(file = file.path(raw_data_path, "Kenya.csv"))
sweden_data <- read.csv(file = file.path(raw_data_path, "Sweden.csv"))
```

The data are collected for a period of 5 years where *person-year* is
a measure of the time contribution of each person during the
period. For example, a person that lives through the entire 5 year
period contributes 5 person-years whereas someone who only lives
through the first half of the period contributes 2.5 person-years.

## Question 1

We begin by computing *crude birth rate* (CBR) for a given
  period.  The CBR is defined as:
  $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
  
  Compute the CBR for each period, separately for Kenya, Sweden, and
  the world.  Start by computing the total person-years, recorded as a
  new variable within each existing `data.frame` via the
  `$` operator, by summing the person-years for men and women.
  Then, store the results as a vector of length 2 (CBRs for two
  periods) for each region with appropriate labels. You may wish to
  create your own function for the purpose of efficient programming.
  Briefly describe patterns you observe in the resulting CBRs.
  
  
  
### Answer 1



```{r}
# Create new variable py = total person years for each data set
world_data$py  <- world_data$py.men  + world_data$py.women
kenya_data$py  <- kenya_data$py.men  + kenya_data$py.women
sweden_data$py <- sweden_data$py.men + sweden_data$py.women
```

<!-- Whenever you find that you will repeat the same calculation -->
<!-- on different sets of data, **consider** creating a function -->
<!-- instead of repeating the code for each piece of data. -->

```{r}
# Function to compute the Crude Birth Rate (CBR)
compute_cbr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
    summarise(cbr = sum(births) / sum(py)) %>% 
    pull()
}
```

<!-- The "pipe" `%>%` from package `magrittr` takes the output -->
<!-- from the left hand side and makes it the first argument of -->
<!-- the expression on the right hand side. -->

<!-- Starting in R 4.1.0, there is a base pipe operator `|>` -->
<!-- which has a similar behavior. -->

```{r}
# Compute the CBR for each data set
(world_cbr  <- compute_cbr(world_data))
(kenya_cbr  <- compute_cbr(kenya_data))
(sweden_cbr <- compute_cbr(sweden_data))
```

The CBR for the world from 1950 to 1955 is greater than its CBR from 2005 to 2010. The CBR for Keyna from 1950 to 1955 is greater than its CBR from 2005 to 2010. The CBR for Sweden from 1950 to 1955 is very slightly greater than its CBR from 2005 to 2010. Kenya has the greatest CBR across itself, Sweden, and the world for both the periods 1950 to 1955 and 2005 to 2010.

  
## Question 2
  
The CBR is easy to understand but contains both men and women
   of all ages in the denominator. We next calculate the *total
     fertility rate* (TFR). Unlike the CBR, the TFR adjusts for age
   compositions in the female population.  To do this, we need to
   first calculate the *age specific fertility rate* (ASFR), which
   represents the fertility rate for women of the reproductive age
   range $[15, 50)$.  The ASFR for age range $[x, x+\delta)$, where
   $x$ is the starting age and $\delta$ is the width of the age range
   (measured in years), is defined as:
  $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$
  Note that square brackets, $[$ and $]$, include the limit whereas
  parentheses, $($ and $)$, exclude it.  For example, $[20, 25)$
  represents the age range that is greater than or equal to 20 years
  old and less than 25 years old.  In typical demographic data, the
  age range $\delta$ is set to 5 years.  Compute the ASFR for Sweden
  and Kenya as well as the entire world for each of the two periods.
  Store the resulting ASFRs separately for each region.  What does the
  pattern of these ASFRs say about reproduction among women in Sweden
  and Kenya?
  

### Answer 2

```{r}
# Function to compute Age specific fertility rate (ASFR)
compute_asfr <- function(population_data) {
  population_data %>% 
    mutate(asfr = births / py.women)
}
```

```{r}
# Compute ASFR for each data set
world_data  <- compute_asfr(world_data)
kenya_data  <- compute_asfr(kenya_data)
sweden_data <- compute_asfr(sweden_data)
```

```{r}
# Compare ASFRs for Kenya and Sweden
kenya_data$asfr
sweden_data$asfr
```

This pattern of ASFRs between Kenya and Sweden suggests that reproduction among women was greater in Kenya between 1950 and 1955 than in between 2005 and 2010, and that reproduction among women was greater in Sweden between 1950 and 1955 than in between 2005 and 2010. Across the two countries, reproduction among women was greater in Kenya than Sweden both between 1950 and 1955 *and* between 2005 and 2010.


## Question 3

Using the ASFR, we can define the TFR as the average number of
  children women give birth to if they live through their entire
  reproductive age.
  $$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$
  
  We multiply each age-specific fertility rate rate by 5 because the
  age range is 5 years. Compute the TFR for Sweden and Kenya as well
  as the entire world for each of the two periods.  As in the previous
  question, continue to assume that women's reproductive age range is
  $[15, 50)$. Store the resulting two TFRs for each country or the
  world as a vector of length two. In general, how has the number of
  women changed in the world from 1950 to 2000? What about the total
  number of births in the world?


### Answer 3

```{r}
# Function to compute the total fertility rate (TFR)
compute_tfr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
    summarise(tfr = 5 * sum(asfr)) %>% 
    pull()
}
```

```{r}
# Compute the TFR for each data set
(world_tfr  <- compute_tfr(world_data))
(kenya_tfr  <- compute_tfr(kenya_data))
(sweden_tfr <- compute_tfr(sweden_data))
```

```{r}
# Compute totals of women and births in the world by period
(world_data %>% 
    group_by(period) %>% 
    summarise(total_women  = sum(py.women),
              total_births = sum(births)) ->
              totals_world)

# Compare how much these totals have changed
(changes_totals <- totals_world[2, -1] / totals_world[1, -1])
```

From 1950 to 2000, both the number of women in the world and the total number of births increased. 


## Question 4

Next, we will examine another important demographic process:
  death. Compute the *crude death rate* (CDR), which is a concept
  analogous to the CBR, for each period and separately for each
  region. Store the resulting CDRs for each country and the world as
  a vector of length two. The CDR is defined as:
  $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$
  Briefly describe patterns you observe in the resulting CDRs.

### Answer 4

```{r}
# Function to compute the Crude death rate (CDR)
compute_cdr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
    summarise(cdr = sum(deaths) / sum(py)) %>% 
    pull()
}
```

```{r}
# Compute the CDR for each data set
(world_cdr  <- compute_cdr(world_data))
(kenya_cdr  <- compute_cdr(kenya_data))
(sweden_cdr <- compute_cdr(sweden_data))
```

The CDR for the world from 1950 to 1955 is greater than its CDR from 2005 to 2010. The CDR for Keyna from 1950 to 1955 is greater than its CDR from 2005 to 2010. The CDR for Sweden from 1950 to 1955 is about the same as its CDR from 2005 to 2010. Kenya has the greatest CDR across itself, Sweden, and the world for both the periods 1950 to 1955 and 2005 to 2010. However, Kenya's CDR for the period 2005 to 2010 is about the same as Sweden's.


## Question 5

One puzzling finding from the previous question is that the CDR
  for Kenya during the period of 2005-2010 is about the same level as
  that for Sweden. We would expect people in developed countries like
  Sweden to have a lower death rate than those in developing countries
  like Kenya. While it is simple and easy to understand, the CDR does
  not take into account the age composition of a population. We
  therefore compute the *age specific death rate* (ASDR). The
  ASDR for age range $[x, x+\delta)$ is defined as:
  $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}
               {\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$
  Calculate the ASDR for each age group, separately for Kenya and
  Sweden, during the period of 2005-2010. Briefly describe the
  pattern you observe.
 
### Answer 5

```{r}
# Function to compute Age specific death rate (ASDR)
compute_asdr <- function(population_data) {
  population_data %>% 
    mutate(asdr = deaths / py)
}
```

```{r}
# Compute ASDR for each data set
world_data  <- compute_asdr(world_data)
kenya_data  <- compute_asdr(kenya_data)
sweden_data <- compute_asdr(sweden_data)
```

During the period of 2005 to 2010, the ASDR is greater in Kenya than in Sweden for each age group.
 
 
## Question 6

One way to understand the difference in the CDR between Kenya
  and Sweden is to compute the counterfactual CDR for Kenya using
  Sweden's population distribution (or vice versa). This can be done
  by applying the following alternative formula for the CDR.
  $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$
  where $P_{[x, x+\delta)}$ is the proportion of the population in the
  age range $[x, x+\delta)$. We compute this as the ratio of
  person-years in that age range relative to the total person-years
  across all age ranges. To conduct this counterfactual analysis, we
  use $\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$
  from Sweden during the period of 2005--2010. That is, first
  calculate the age-specific population proportions for Sweden and
  then use them to compute the counterfactual CDR for Kenya. How does
  this counterfactual CDR compare with the original CDR of Kenya?
  Briefly interpret the result.

### Answer 6

```{r}
# Function to compute population proportion by period
compute_pop_prop <- function(population_data) {
  population_data %>% 
    group_by(period) %>%
    mutate(pop_prop = py / sum(py)) %>%
    ungroup()
}
```

```{r}
# Compute population proportion for each data set
world_data  <- compute_pop_prop(world_data)
kenya_data  <- compute_pop_prop(kenya_data)
sweden_data <- compute_pop_prop(sweden_data)
```

```{r}
# Compute Kenyas CDR Kenya had Sweden's population distribution
mutate(kenya_data,
       temp_cdr = asdr * sweden_data$pop_prop) %>%
       group_by(period) %>%
       summarise(cdrresweden = sum(temp_cdr))
```

During the period of 1950 to 1955, Kenya's counterfactual CDR is about the same as its original CDR. However, during the period of 2005 to 2010, Kenya's counterfactual CDR is greater than its original CDR; this suggests that, had Kenya had Sweden's population distribution at the time of measurement, Kenya's CDR would be considered more significant when compared with using its own population distribution.